---
title: "Primary Feather Evolution"
author: "Angry_biRds"
date: "2022-12-15"
output: html_document
bibliography: feathers.bib
---

```{r, echo= FALSE, include= FALSE}
library(tidyverse)
library(Momocs)
library(vroom)
library(ape)
library(mgsub)
library(phytools)
library(RRphylo)
library(ggtree)
library(wesanderson)
```

### Introduction
Migration is an integral facet of many avian life cycles and usually leads birds from areas of low resources to areas of high resources and nesting spaces. To effectively move between distant geographic locations, migratory birds have been naturally selected for specific morphological traits, such as the length/size of wings and feathers. Most recent research has focused on the overall length of the wing and its outline in the study of the external avian morphological adaptations as it can facilitate an energy-efficient migratory flight [@de2020mechanical]. Primary feathers are a vital part of a bird’s wing and overall flight ability. These feathers are one of the most prominent features found on a bird’s wings, generating the forward thrust during flight which ultimately aids in propelling the bird into the air. If these feathers are cut or damaged, the bird’s ability to fly is significantly impaired. 

In this study, we will explore primary feather length, shape, and structural characteristics that may explain the variation between varying species of birds, both migratory and non-migratory. We predict that migration distance will be strongly correlated with longer primary feathers. These mechanical and structural adaptations within the primary feathers may give insight into the importance/favoring of primary feathers in the long-distance flight of birds. This study also aims to determine if the rates of primary feather-shape evolution between the two types of birds are similar and if they have undergone significant shifts among lineages. The feathers will be evaluated and compared through the phylogenetic analysis of over 100 species of Aves within different families, utilizing data collected from the feather atlas provided by the U.S. Fish & Wildlife Services. This research is vital as avian migration is an ever-changing process, and scientists predict that migration patterns will significantly shift in the distant future as the planet continues to warm due to climate change. 

### Methods
#### Image Acquisition 
Images analyzed in this project were acquired through the U.S. Fish & Wildlife Service’s database: “Feather Atlas”. Avian families analyzed include Apodiformes, Colmbiformes, Galliformes, Gruiformes, & Passeriformes. Records from the Feather Atlas are accessible for all Aves further analyzed below. 

#### Digitization
The outermost primary feather shapes across all species of Avians were examined utilizing the image analysis program FIJI. After opening each image in FIJI, the scale was set aligning with the scale-bar present in the database. The primary feathers were then traced utilizing the polygon selection tool. A script was created, and the selected points (x and y coordinates) were exported in a .txt file for further analysis.

#### Shape & Comparative Analysis
Outline-based shape analysis, utilizing Elliptical Fourier Analysis (EFA) and Principal Components Analysis (PCA), were performed to compare shape change between species. 
PCA plot output for primary feathers can be referenced below (Figures 2 & 3). Evolutionary rates of the PC scores were observed across the Avian tree [@BirdTree.org] and species information was added (Figure 1). 

#### Evolutionary Rates & Shifts
The “non-censored test” [@o2006testing] was utilized to estimate the rates of morphological evolution over a phylogenetic tree. Phylogenetic ridge regression [@kratsch2014ridgerace] was then utilized to determine whether the evolutionary shift was to a slower or faster rate. The regressions are based on the phenotype values at each node vs. time between nodes and along the path. 


```{r, echo= FALSE, message= FALSE, fig.cap= "Figure 1: Modified phylogenetic tree sourced from @BirdTree.org. The tree was shortened to only include species analyzed in PCA."}

con.tree <- read.tree("contree.tre")
con.tree <- root(con.tree,"Sayornis_phoebe")
plot(main = "Birds Phylogenetics Tree", con.tree,cex=0.2)

```


### Results

#### Principal Components

```{r, echo= FALSE, include= FALSE}

f <- list.files("Feathers",pattern=".txt|.csv",full.names = TRUE)

out.df <- vroom::vroom(f, id = "filename")
#make list
outs.l <- sapply(f,function(x) out.df %>% filter(filename==x) %>% select(X,Y) %>% as.matrix)
summary(outs.l)

birdlist <- tibble(xy.file=unique(basename(out.df$filename)), species=unique(mgsub(basename(out.df$filename), c("XY_",".txt"), c("",""))))

outs.l %>% 
  Out() %>% 
  coo_flipx()

##

#make a large df with vroom
out.df <- vroom::vroom(f, id = "filename") %>% 
  mutate(species=gsub("XY_.+_\\..+","\\1",basename(filename))) %>% 
  na.omit()

#make list
outs.l <- sapply(f,function(x) out.df %>% filter(filename==x) %>% select(X,Y) %>% as.matrix)

#extract wing info
species <- gsub("XY_.+_\\..+","\\1",basename(names(outs.l)))


outs <-  outs.l %>% 
  Out(fac=list(species=species)) %>% 
  coo_flipx()
#pipes the list of matrices into Our() function, flip, and visualize with stack

outs.min <- outs %>% 
  coo_nb() %>% 
  min()

outs %>% 
  coo_interpolate(outs.min) %>% 
  coo_slide(id=1) %>% 
  coo_align()  %>%
  fgProcrustes()

outs.pca <- outs %>%
  coo_interpolate(outs.min) %>%
  coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  efourier(norm=FALSE) %>% 
  PCA()
```

```{r, echo= FALSE, message= FALSE, fig.cap= "Figure 2: Biplot visualization of Principle Component 1 (PC1) and PC2 of the primary feathers."}
outs.pca %>% 
  plot_PCA(title = "Primary Feathers")
```

Based on the PCA analysis results shown in figure 2, PC1, diagonal lobe thickness, and PC2, equatorial rotation, accounts for 63.5% and 24.9% of the variance in our sample respectively. 



```{r, echo= FALSE, include= FALSE}
outs.pca <-  tibble(xy.file=basename(rownames(outs.pca$x)),PC1=outs.pca$x[,1],PC2=outs.pca$x[,2])%>% 
  left_join(birdlist)

#PC1s
outs.pc1 <- outs.pca %>% 
  filter(species%in% con.tree$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull

names(outs.pc1) <-  outs.pca%>% 
  filter(species%in% con.tree$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)


#PC2s
outs.pc2 <- outs.pca %>% 
  filter(species%in% con.tree$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(outs.pc2) <-  outs.pca%>% 
  filter(species%in% con.tree$tip.label) %>% 
  group_by(species) %>%
  summarize(PC2=mean(PC2)) %>% 
  pull(species)

outs.PC1.BM<-brownie.lite(con.tree,outs.pc1*10)

outs.PC2.BM<-brownie.lite(con.tree,outs.pc2*10)

outs.PC1.RR <- RRphylo(tree=con.tree,y=outs.pc1)
outs.PC2.RR <- RRphylo(tree=con.tree,y=outs.pc2)

outs.PC1.SS<- search.shift(RR=outs.PC1.RR,status.type="clade")
outs.PC2.SS<- search.shift(RR=outs.PC2.RR,status.type="clade")
```

<br>

#### Evolutionary Rates: PC1 vs PC2

##### Estimated Evolutionary Rate of PC1
```{r, echo= FALSE}
outs.PC1.BM$sig2.single
```

##### Estimated Evolutionary Rate of PC2
```{r, echo= FALSE}
outs.PC2.BM$sig2.single
```

Across all observed primary feathers of various bird species, the evolutionary rate of PC1 is 0.006242593 and that of PC2 is 0.001396271, meaning PC1 is evolving at more than 4 times the rate of PC2. This could indicate there has been an elevated amount of selection pressure for PC1 in primary feathers as a potential advantageous trait in bird flight mechanics in varying environments and lifestyles when compared to other principal components. 

<br>

#### Evolutionary Rate Shifts

##### PC1

```{r, echo=FALSE, message=FALSE, fig.cap= "Figure 3: Evolutionary rate shift in the primary feathers' PC1 plotted on modified phylogenetic tree. The numeric label represents at which node in the tree the shift occurred at. The table below shows the magnitude and statistical significance of the rate difference compared to the rest of the tree."}
plot(con.tree, cex=0.1)
nodelabels(node = as.numeric(rownames(outs.PC1.SS$single.clades)),text = rownames(outs.PC1.SS$single.clades))

```

```{r, echo= FALSE}
outs.PC1.SS$single.clades
```

According to figure 3 and the table, there is a statistically significant downwards PC1 rate shift in node 172, which corresponds to the fringillidae family. Compared to other groups, the fringillidae's primary feather PC1 evolutionary rate is decreased by a magnitude of 0.0009462863 at a 0.1% significance level. 

##### PC2

```{r, echo=FALSE, message=FALSE, fig.cap= "Figure 4: Evolutionary rate shift in the primary feathers' PC2 plotted on modified phylogenetic tree. The table below shows the magnitude and statistical significance of the rate difference compared to the rest of the tree."}
plot(con.tree, cex=0.1)
nodelabels(node = as.numeric(rownames(outs.PC2.SS$single.clades)),text = rownames(outs.PC2.SS$single.clades))
```

```{r, echo= FALSE}
outs.PC2.SS$single.clades
```

Similarly, there is also a statistically significant downwards PC2 rate shift in node 175, which represents the group encompassing the new world blackbirds and the new world sparrows. Within this group, the evolutionary rate of PC2 in primary feathers is shifted by -0.0003019869 at a 0.4% significance level. Based on these results regarding the outermost primary feathers, which are used for thrust and lift during flight, we can speculate that certain groups have evolved varying degrees of traits that may affect their flight mechanisms and potentially behaviors such as migration.

### Discussion
Since the groups in which we detected significant evolutionary rate shifts broadly include both migrant and resident species, the shifts did not track with migrational birds. The double downward shift, with PC2 downshifting a large portion of measured birds and PC1 further downshifting some of the birds, supports the theory that primary flight feathers have a slow evolutionary rate. This slow evolutionary rate is likely because of the difficulty to evolve flight, and the considerable selective advantage efficient flight is on bird species [@Bock]. Changes to the feather are less common, as any change that decreases flight efficiency often proves to be much more detrimental to the bird than any gained benefit. The modern bird lineages studied likely had an ancestor that already had a highly adapted feather shape that worked well for both migratory and nonmigratory behavior. It is interesting to note that these down shifts occurred within New World lineages, suggesting a geographical connection to evolutionary rate.

To continue this study, all of the primaries of the birds should be analyzed so that the feathers' evolutionary rate can be compared to one another. Additionally, more bird lineages should be studied to allow for a broader understanding of primary feather shape, and details about their habits (beyond migratory or nonmigratory) should be tracked to see if other factors are linked with changes in feather shape. Furthermore, previous studies have been conducted regarding the feather quality of long-distance migratory specie [@szép_dobránszky_møller_dyke_lendvai_2019]. Researchers utilized measurements of bending stiffness, feather length, and thickness to lead them to the conclusion that older individuals tend to posses better quality feathers. They also concluded that individuals who live to older ages have similar, or higher, feather growth rates than those captured at young ages. Relating to our research, utilizing known age feather samples, instead of the generalized Adult group, could provide further information regarding the adaptive significance of the traits showcased in the aforementioned study and specifically within each of the clades analyzed in our research.

### Contributions
Philip Chen (Code + Results + Data Collection)

Stephen Mooney (Code + Discussion)

Jack Kloster (Introduction + Methods + Data Collection)

Jordan Nunes (Data Collection + Discussion)


### CPK comments

Just excellent work, guys!! I'm really impressed how you conceived and executed this project. It's an interesting question that was answered by leveraging what you learned in the class. Congrats!!! 150/150

### References